Evolution of precopulatory and postcopulatory inbreeding behaviour

### 2016-09-12 ms_pre_post.Rmd
Tweak Model section

We track interactions among discrete individuals in a small focal population that persists over multiple generations (key individual traits, parameter values, and variables are described in Table 1). Individuals within this population are either female or male, and can be related to each other if they share a common ancestor. We record the coefficient of kinship ($k$), defined as the probability that a randomly selected homologous allele from each individual is identical-by-descent [@Lynch1998], between all pairs of individuals in each generation using the population pedigree [see @Boyce1983; @Duthie]. At the start of each generation, each female pays up to three independent costs of (1) tendency to engage in polyandry ($c_{P}$), (2) mating strategy with respect to inbreeding ($c_{M}$), and (3) fertilisation strategy with respect to inbreeding ($c_{F}$). The number of males that a focal female mates with ($N_{males}$) is determined by the effects of 20 alleles ($P_{a}$) across 10 diploid loci that combine additively to determine her genotype ($P_{g}$), and ultimately her phenotype for tendency to engage in polandry ($P_{p}$). A focal female's choice of mate(s) is affected by 20 alleles ($M_{a}$) across 10 diploid loci that combine additively to affect her genotype ($M_{g}$) and phenotype ($M_{p}$) for pre-copulatory inbreeding avoidance or preference, and kinship between the focal female and her potential mates. Similarly, after mating has occurred, post-copulatory fertilisation of a focal female's $n$ total offspring is affected by 20 alleles ($F_{a}$) across 10 diploid loci that combine additively to affect the genotype ($F_{g}$) and phenotype ($F_{p}$) of the degree to which fertilisation success is affected by kinship. After fertilisation, the viability of a focal female's $n$ offspring is decreased by inbreeding depression ($\beta$), and surviving offspring become adults. After inbreeding depression in offspring, $\rho$ adult male immigrants arrive in the population, then density effects reduce the entire population down to female ($K_{f}$) and male ($K_{m}$) carrying capacity. Over generations, we record mean values of alleles underlying polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy.

@@

Each female chooses a number of males to mate with ($N_{males}$) based on her genetically determined tendency to engage in polyandry ($P_{p}$). The exact value of $N_{males}$ is calculated by randomly sampling from a poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This random sampling procedure causes some stochastic variation in the number of realised mates a that female chooses, with a mean $N_{males}$ of $P_{p}+1$. A value of one is added to $P_{p}$ to ensure that all females choose at least one mate, as low values of $P_{p}$ might return $Poisson(P_{p})$ values of zero.

The value of $P_{p}$ is affected by the summed values of her 20 $P_{a}$ alleles. Inheritence of all alleles models a diploid genetic architecture in which 10 $P_{a}$ alleles are randomly sampled from each parent with no physical linkage. Values of $P_{a}$ can take any real number [continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014]. The tendency for polyandry genotype $P_{g}$ is defined by the sum of all of a focal female's 20 $P_{a}$ values, which might be negative or positive. However, the tendency for polyandry phenotype $P_{p}$ is restricted to zero or positive values because the number of males that a female chooses to mate with cannot be negative. Hence, negative values of $P_{g}$ are translated to $P_{p}$ values of zero such that $P_{p} = 0$ when $P_{g} < 0$, but when $P_{g} \geq 0$, then $P_{p} = P_{g}$ (i.e., the phenotype value of polyandry equals the genotype value of polyandry if and only if the genotype value is greater than or equal to zero, else the phenotype value of polyandry always equals zero). The resulting $P_{p}$ is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998], in which polyandry is expressed discretely at the threshold value of $P_{g} > 0$, but is determined by the consequence of continuous genetic variation arising from the effects of $P_{a}$ alleles that each contribute additively to a female's tendency toward polyandry. Negative genotype values of $P_{g}$ therefore all have the same phenotype, resulting in consistently monandrous females. In contrast, positive values of $P_{g}$ and therefore $P_{p}$ can lead to different degrees of realised polyandry in the form of discrete numbers of mates ($N_{males}$).

@@

Alleles of a focal female's offspring mutate with a probability of $\mu=0.001$, with mutation being an independent event the does or does not occur for each locus. In general, loss of function mutations might be expected to be more frequent than mutations that increase the functionality of adaptations that rely on complex physiologies or behaviours. If tendency for polyandry, inbreeding mating strategies, or inbreeding fertilisation strategies rely on such complex physiologies or behaviours, then we might expect the direction of mutation to be biased (e.g., towards monandry or random mating). However, because we are interested specifically in how selection affects reproductive strategy given varying costs and inbreeding depression, we model mutation as unbiased so that selection can be isolated and inferred clearly from mean allele values.

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a function of the kinship coefficient between $i$ and the sire of her offspring $j$ ($k_{ij}$), and a log-linear slope of inbreeding depression $\beta$. We model selection as having an absolute effect on offspring viability (i.e., hard selection), rather than a relative effect on viability (i.e., soft selection), so that the effect of $\beta$ is consistent over generations and across different parameter combinations. Natural selection on tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$) can therefore be compared with respect to log-linear inbreeding depression slopes across generations and parameter combinations. If offspring viability was instead relative such that the least inbred offspring had a high viability regardless of the absolute value of $k_{ij}$, then the effect of kinship coefficients and inbreeding depression slopes on offspring viability would depend on the population-wide distribution of $k_{ij}$ between females and males, making model inference more difficult. Therefore, we define offspring viability as,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
The slope $\beta$ models the number of haploid lethal equivalents that exist in the gametes of $i$ and $j$, and which might combine to become homozygous in offspring and reduce viability. Equation 1 assumes that each deleterious recessive allele affects offspring viability independently, meaning that every deleterious recessive allele added will have a multiplicative affect on decreasing total viability [@Morton1956; @Mills1994], causing the log-linear relationship between parent kinship and inbreeding depression. We assume that inbreeding depression always decreases offspring viability, $\beta \geq 0$ (i.e., no outbreeding depression). Therefore, because $k_{ij}$ is constrained to values between zero and one, the product of $-\beta$ and $k_{ij}$ must be a value less than or equal to zero, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We can therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability using a Bernoulli trial.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different parameter combinations test our four focal hypotheses (see below). We simulate up to eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be a strong cost of tendency for polyandry] for values of $\beta = 3$ ($8$ total combinations). In all simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each combination of parameter values simulated is replicated $40$ times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allowed us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 10000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we also set the number of immigrants per generation to $\rho=5$, which produced a range of kin and non-kin in each generation such that females could express inbreeding strategies. Females always produced $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Carrying capacities for females ($K_{f}$) and males ($K_{m}$) were both set to 100 because previous modelling has shown that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).

To test our hypothesis that selection for evolution of polyandry is driven by female's ability to bias fertilisation toward inbreeding avoidance through post-copulatory mechanisms, we compare simulations in which all three phenotypes including tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$) evolve, with an identical set of simulations in which $F_{p}$ cannot evolve. In simulations without evolving $F_{p}$, we therefore include all $4$ combinations of $c_{P}=\{0,0.02\}$ and $c_{M}=\{0,0.02\}$. To stop $F_{p}$ from evolving, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes resulting from $F_{a}$ cause random fertilisation with respect to kinship (i.e, $F_{p}=0$). Hence, while mutation and drift can still occur for $F_{a}$ alleles, these alleles have no phenotypic effect and polyandry cannot cause polyandrous females to benefit through an increased ability to bias fertilisation toward inbreeding avoidance (under such conditions, there can also never be a cost of $F_{p}$, meaning that $c_{F}=0$). By comparing evolution of mean $P_{p}$ values over generations when $F_{p}$ also evolves versus when $F_{p}=0$ (random fertilisation), we test whether or not evolution of inbreeding fertilisation strategy increases selection for evolution of polyandry.

To test hypotheses regarding evolution under different relative costs of phenotypes, we consider all combinations of costs using simulations in which all three traits evolve and $\beta=3$. To test our hypothesis that inbreeding mating strategy will evolve toward inbreeding avoidance ($M_{p}<0$) when mating costs are low relative to the costs of tendency for polyandry and inbreeding fertilisation strategy, but that selection for inbreeding fertilisation strategy will then be negligible ($F_{p} \approx 0$), we quantify evolution of $M_{p}$ and $F_{p}$ over generations in simulations where $c_{M}=0$ but $c_{P}$, $c_{F}$, or both equals $0.02$. We compare evolution in these three cost parameter value combinations with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $M_{p}$ but not $F_{p}$ will evolve to more negative values when $c_{M}$ is relatively low than when all cost values are uniform. To test the hypothesis that inbreeding fertilisation strategy will evolve toward inbreeding avoidance ($F_{p}<0$) given low costs of polyandry and inbreeding fertilisation strategy relative to the cost of inbreeding mating strategy, but that selection on inbreeding mating strategy will be negligible ($M_{p} \approx 0$), we quantify evolution of $M_{p}$ and $F_{p}$ over generations when $c_{F}=0$, $c_{P}=0$, and $c_{M}=0.02$ . We compare evolution for this combination of cost parameter values with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $F_{p}$ but not $M_{p}$ will evolve to more negative values when $c_{F}$ is relatively low than when all cost values are uniform.

>>>>

We track interactions among discrete individuals in a small focal population that persists over multiple generations (key individual traits, parameter values, and variables are described in Table 1). Individuals within this population are either female or male, and can be related to each other if they share a common ancestor. We record the coefficient of kinship ($k$), defined as the probability that a randomly selected homologous allele from each individual is identical-by-descent [@Lynch1998], between all pairs of individuals in each generation using the population pedigree [see @Boyce1983; @Duthie]. At the start of each generation, each female pays up to three independent costs of (1) tendency to engage in polyandry ($c_{P}$), (2) mating strategy with respect to inbreeding ($c_{M}$), and (3) fertilisation strategy with respect to inbreeding ($c_{F}$). The number of males that a focal female mates with ($N_{males}$) is determined by the effects of 20 alleles ($P_{a}$) across 10 diploid loci that combine additively to determine her genotype ($P_{g}$), and ultimately her phenotype for tendency to engage in polandry ($P_{p}$). A focal female's choice of mate(s) is affected by 20 alleles ($M_{a}$) across 10 diploid loci that combine additively to affect her genotype ($M_{g}$) and phenotype ($M_{p}$) for pre-copulatory inbreeding avoidance or preference, and kinship between the focal female and her potential mates. Similarly, after mating has occurred, post-copulatory fertilisation of a focal female's $n$ total offspring is affected by 20 alleles ($F_{a}$) across 10 diploid loci that combine additively to affect the genotype ($F_{g}$) and phenotype ($F_{p}$) of the degree to which fertilisation success is affected by kinship. After fertilisation, the viability of a focal female's $n$ offspring is decreased by inbreeding depression ($\beta$), and surviving offspring become adults. After inbreeding depression in offspring viability, $\rho$ adult male immigrants arrive in the population, then density effects reduce the entire population down to separate female ($K_{f}$) and male ($K_{m}$) carrying capacities Over generations, we record mean values of alleles underlying polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy.

@@

After costs are paid, each remaining female chooses a number of males to mate with ($N_{males}$) based on her genetically determined tendency to engage in polyandry ($P_{p}$). The exact value of $N_{males}$ is calculated by randomly sampling from a poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This random sampling procedure causes some stochastic variation in the number of realised mates a that female chooses, with a mean $N_{males}$ of $P_{p}+1$. A value of one is added to $P_{p}$ to ensure that all females that have not paid a cost choose at least one mate, as low values of $P_{p}$ might return $Poisson(P_{p})$ values of zero.

The value of $P_{p}$ is affected by the summed values of a focal female's 20 $P_{a}$ alleles. Inheritence of all alleles models a diploid genetic architecture in which 10 $P_{a}$ alleles are randomly sampled from each parent with no physical linkage. Values of $P_{a}$ can take any real number [continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014]. The tendency for polyandry genotype $P_{g}$ is therefore defined by the sum of all of a focal female's 20 $P_{a}$ values, which might be negative or positive. However, the tendency for polyandry phenotype $P_{p}$ is restricted to zero or positive values because the number of males that a female chooses to mate with cannot be negative. Hence, negative values of $P_{g}$ are translated to $P_{p}$ values of zero such that $P_{p} = 0$ when $P_{g} < 0$, but when $P_{g} \geq 0$, then $P_{p} = P_{g}$ (i.e., the phenotype value of polyandry equals the genotype value of polyandry if and only if the genotype value is greater than or equal to zero, else the phenotype value of polyandry always equals zero). The resulting $P_{p}$ is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998], in which polyandry is expressed discretely at the threshold value of $P_{g} > 0$, but is determined by the consequence of continuous genetic variation arising from the effects of $P_{a}$ alleles that each contribute additively to a female's tendency toward polyandry. Negative genotype values of $P_{g}$ therefore all have the same phenotype, resulting in consistently monandrous females. In contrast, positive values of $P_{g}$ and therefore $P_{p}$ can lead to different degrees of realised polyandry in the form of discrete numbers of mates ($N_{males}$).

@@

Alleles of a focal female's offspring mutate with a probability of $\mu=0.001$, with mutation being an independent event that does or does not occur for each locus. In general, loss of function mutations might be expected to be more frequent than mutations that increase the functionality of adaptations that rely on complex physiologies or behaviours. If tendency for polyandry, inbreeding mating strategies, or inbreeding fertilisation strategies rely on such complex physiologies or behaviours, then we might expect the direction of mutation to be biased (e.g., towards monandry or random mating). However, because we are interested specifically in how selection affects reproductive strategy given varying costs and inbreeding depression, we model mutation as unbiased so that selection can be isolated and inferred clearly from mean allele values.

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a function of the coefficient of kinship between $i$ and the sire of her offspring $j$ ($k_{ij}$), and a log-linear slope of inbreeding depression $\beta$. We model selection as having an absolute effect on offspring viability (i.e., hard selection), rather than a relative effect on viability (i.e., soft selection), so that the effect of $\beta$ is consistent over generations and across different parameter combinations. Natural selection on tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$) can therefore be compared across generations and parameter combinations. If offspring viability was instead relative such that the least inbred offspring had a high viability regardless of the absolute value of $k_{ij}$, then the effect of kinship coefficients and inbreeding depression slopes on offspring viability would depend on the population-wide distribution of $k_{ij}$ between females and males, making model inference more difficult. Therefore, we define offspring viability as,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
The slope $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might combine to become homozygous in offspring and reduce viability. Equation 1 assumes that each deleterious recessive allele affects offspring viability independently, meaning that every deleterious recessive allele added will have a multiplicative affect on decreasing total viability [@Morton1956; @Mills1994], causing the log-linear relationship between parent kinship and inbreeding depression. We assume that inbreeding depression always decreases offspring viability, $\beta > 0$ (i.e., no outbreeding depression). Therefore, because $k_{ij}$ is constrained to values between zero and one, the product of $-\beta$ and $k_{ij}$ must be a value less than or equal to zero, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We can therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability using a Bernoulli trial.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different parameter combinations to test our four focal hypotheses (see below). We simulate up to eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be a strong cost of tendency for polyandry] for values of $\beta = 3$ ($8$ total combinations). In all simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each combination of parameter values simulated is replicated $40$ times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allow us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 10000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we also set the number of immigrants per generation to $\rho=5$, which produced a range of kin and non-kin in each generation allowing females could express inbreeding strategies. Females always produced $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Carrying capacities for females ($K_{f}$) and males ($K_{m}$) were both set to 100 because previous modelling has shown that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).

To test our hypothesis that selection for evolution of polyandry is driven by a female's ability to bias fertilisation toward inbreeding avoidance through post-copulatory mechanisms, we compare simulations in which all three phenotypes including tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$) evolve, with an identical set of simulations in which $F_{p}$ cannot evolve. In simulations without evolving $F_{p}$, we therefore include all $4$ combinations of $c_{P}=\{0,0.02\}$ and $c_{M}=\{0,0.02\}$ (but not $c{F}$). To stop $F_{p}$ from evolving, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes resulting from $F_{a}$ cause random fertilisation with respect to kinship (i.e, $F_{p}=0$). Hence, while mutation and drift can still occur for $F_{a}$ alleles, these alleles have no phenotypic effect and polyandry cannot cause polyandrous females to benefit through an increased ability to bias fertilisation toward inbreeding avoidance (under such conditions, there can also never be a cost of $F_{p}$, meaning that $c_{F}=0$). By comparing evolution of mean $P_{p}$ values over generations when $F_{p}$ also evolves versus when $F_{p}=0$ (random fertilisation), we test whether or not evolution of inbreeding fertilisation strategy increases selection for evolution of polyandry.

To test hypotheses regarding evolution under different relative costs of phenotypes, we consider all combinations of costs using simulations in which all three traits evolve. To test the hypothesis that inbreeding mating strategy will evolve toward inbreeding avoidance ($M_{p}<0$) when mating costs are low relative to the costs of tendency for polyandry and inbreeding fertilisation strategy, but that selection for inbreeding fertilisation strategy will then be negligible ($F_{p} \approx 0$), we quantify change in $M_{a}$ and $F_{a}$ over generations in simulations where $c_{M}=0$ but $c_{P}$, $c_{F}$, or both equals $0.02$. We compare evolution in these three cost parameter value combinations with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $M_{p}$ but not $F_{p}$ will evolve to more negative values when $c_{M}$ is relatively low than when all cost values are uniform. To test the hypothesis that inbreeding fertilisation strategy will evolve toward inbreeding avoidance ($F_{p}<0$) given low costs of polyandry and inbreeding fertilisation strategy relative to the cost of inbreeding mating strategy, but that selection on inbreeding mating strategy will be negligible ($M_{p} \approx 0$), we quantify evolution of $M_{p}$ and $F_{p}$ over generations when $c_{F}=0$, $c_{P}=0$, and $c_{M}=0.02$ . We compare evolution for this combination of cost parameter values with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $F_{a}$ but not $M_{a}$ allele values will become more negative over generations when $c_{F}$ is relatively low than when all cost values are uniform.
